Forcing for Isabelle

### 2018-02-20 blueprint.tex
first proof, Strengthening Lemma.

The  function $F$ will essentially have 
the same  arguments  of $\forceisa$, but we will deconstruct the
formula and pack everything as a tuple. Namely, we will have
\begin{align}
  F(\lb\equ,t_1,t_2,\PP,\leq,\1,p\rb) =1 &\iff  
  \forceisa(t_1=t_2,\PP,\leq,\1,p) \label{eq:forcing-F-equ}\\
  F(\lb\bel,s,t_1,\PP,\leq,\1,p\rb) =1 &\iff  
  \forceisa(s\in t_1,\PP,\leq,\1,q). \label{eq:forcing-F-bel}
\end{align}
Here, $\equ$ and $\bel$ are arbitrary sets that encode which type of
formula we are considering. We can choose our $A$ to be
$\{\equ,\bel\}\times V^6$ (i.e., all 7-tuples whose first coordinate
are  $\equ$ or $\bel$).

@@

$G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =1$ if one of the following hold:
\begin{itemize}
\item $\atr=\equ$ and   $\forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :
  f(\lb\bel,s,t_1,\PP,\leq,\1,q\rb) =1  \lsii
  f(\lb\bel,s,t_2,\PP,\leq,\1,q\rb) =1$; or
\item $\atr=\bel$ and $\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  f(\lb\equ,t_1,s,\PP,\leq,\1,q\rb) =1)\}$ is dense below $p$.
\end{itemize}
In any other case, $G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =0$. In the
above, $f(v)=1$ should be interpreted as $\lb v,1\rb\in f$ (i.e., $f$
must be defined for the argument at hand). From this scheme an element
of $\formula$ must be produced.

@@

With these elements we might reproduce the formula $\psi$ that defines
$F$ as in \cite[p.~49]{kunen2011set}. Let  $\app(d,h)$  be the formula
that states:
\begin{enumerate}
\item \label{item:dom-R-cerrado} $d$ is ``downwards $R$-closed'':
  $\forall y \in d, 
  R^{-1}\{y\}\subseteq d$;
\item \label{item:4} $h$ is a function with $\dom h =d \subseteq A$; and
\item \label{item:3} for all $y\in d$, $h(y) = G(y,h\restr(R^{-1}\{y\}))$.
\end{enumerate}
Then the formula defining $F$ is the following:
\[
(x\notin A \y y=0) \o \bigl(x\in A \y \exists d,h\, (\app(d,h)\y
x\in d \y h(x)=y)\bigr).
\]
The well-foundedness of $R$ is used to show that this formula is indeed
functional, and that it satisfies the required recursive equation
involving $G$. Once this is all translated to an element of \formula,
we'll be able to write the base case of the recursive definition of
$\forceisa$ using Equations~(\ref{eq:forcing-F-equ})
and~(\ref{eq:forcing-F-bel}).

@@

\section{The fundamental theorems}
These are stated as two lemmas: \emph{Truth} and
\emph{Definability}. The second one ``simply'' says that the forcing
relation is definable. The previous recursive construction amounts to
that. The Truth Lemma states that the forcing relation indeed relates
satisfaction in $M[G]$ to that in $M$. 
\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
  language of set 
  theory with all free variables displayed. Then the
  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
  \begin{enumerate}
  \item     $M[G]\models\phi(\val(\tau_1,G),\dots,\val(\tau_n,G))$ for
    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
  \item $M\models\forceisa(\phi(\tau_1,\dots,\tau_n),\PP,\leq,\1,p)$.
  \end{enumerate}
\end{lemma}
A bit more formally, one should write the parameters to the left of
$\models$:
\begin{gather*}
  \forall G : p\in G \limp M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
  \models\phi(x_1,\dots,x_n) \\
  \text{if and only if}\\
  M,[\tau_1,\dots,\tau_n,\PP,\leq,\1,p] 
  \models\forceisa(\phi(x_1,\dots,x_n),x_{n+1},x_{n+2},x_{n+3},x_{n+4})
\end{gather*}
The proof of the Truth Lemma is by recursion in elements of
$\formula$. The following auxiliary results (adapted from
\cite[IV.2.43]{kunen2011set}) is also proved by recursion in
$\formula$.
\begin{lemma}\label{lem:strengthen} 
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
  $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
  implies $M\models\forceisa(\phi,\PP,\leq,\1,p_1)$.
\end{lemma}
\begin{lemma}\label{lem:density}
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ if and only if
    $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
    dense below $p$.''
\end{lemma}
\begin{lemma}\label{lem:forcing-negation}
  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
  $M\models\forceisa(\phi,\dots)$ if and only if 
  $M\models\neg\exists
  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
\end{lemma}

\subsection{Truth for atomic formulas}
We start by proving  Lemma~\ref{lem:strengthen}.% and~\ref{lem:density}.
\begin{proof}[Proof of  Lemma~\ref{lem:strengthen} (atomic formulas)]
  Let $M$, $\lb\PP,\leq,\1\rb$, $p\in\PP$ as in the hypothesis of
  Lemma~\ref{lem:truth-lemma}, let $p_1\leq p$ and let $\phi(x_1,x_2)$
  be atomic. We 
  have to show that for every    $\tau_1,\tau_2\in M$,
  \begin{equation} \label{eq:M-phi-atom}
    M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
    \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
  \end{equation}
  then the same holds for the environment
  $[\tau_1,\tau_2,\PP,\leq,\1,q]$.
\end{proof}

>>>>

The  function $F$ will essentially have 
the same  arguments  of $\forceisa$, but we will deconstruct the
formula and pack everything as a tuple. Namely, we will have
\begin{align}
  F(\lb\equ,t_1,t_2,\PP,\leq,\1,p\rb) =1 &\iff  
  \forceisa(t_1=t_2,\PP,\leq,\1,p) \label{eq:forcing-F-equ}\\
  F(\lb\bel,s,t_1,\PP,\leq,\1,p\rb) =1 &\iff  
  \forceisa(s\in t_1,\PP,\leq,\1,q). \label{eq:forcing-F-bel}
\end{align}
Here, $\equ$ and $\bel$ are arbitrary definable sets that encode which type of
formula we are considering (for instance, $0$ and $1$, 
respectively). We can choose our $A$ to be 
$\{\equ,\bel\}\times V^6$ (i.e., all 7-tuples whose first coordinate
are  $\equ$ or $\bel$).

@@

$G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =1$ if one of the following hold:
\begin{itemize}
\item $\atr=\equ$ and   $\forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :
  f(\lb\bel,s,t_1,\PP,\leq,\1,q\rb) =1  \lsii
  f(\lb\bel,s,t_2,\PP,\leq,\1,q\rb) =1$; or
\item $\atr=\bel$ and $\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  f(\lb\equ,t_1,s,\PP,\leq,\1,q\rb) =1)\}$ is dense below $p$.
\end{itemize}
In any other case, $G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =0$. In the
above, $f(v)=1$ should be interpreted as $\lb v,1\rb\in f$ (i.e., $f$
must be defined for the argument at hand). From this scheme an element
of $\formula$ must be produced. (This is the reason why  $\equ,\bel$
were chosen to be definable.)

@@

With these elements we might reproduce the formula $\psi$ that defines
$F$ as in \cite[p.~49]{kunen2011set}. Let  $\app(d,h)$  be the formula
that states:
\begin{enumerate}
\item \label{item:dom-R-cerrado} $d$ is ``downwards $R$-closed'':
  $\forall y \in d, 
  R^{-1}\{y\}\subseteq d$;
\item \label{item:4} $h$ is a function with $\dom h =d \subseteq A$; and
\item \label{item:3} for all $y\in d$, $h(y) = G(y,h\restr(R^{-1}\{y\}))$.
\end{enumerate}
Then the formula defining $F$ is the following:
\[
(x\notin A \y y=0) \o \bigl(x\in A \y \exists d,h\, (\app(d,h)\y
x\in d \y h(x)=y)\bigr).
\]
The well-foundedness of $R$ is used to show that this formula is indeed
functional, and that it satisfies the required recursive equation
involving $G$. The following intermediate result is needed in the
sequel.
\begin{lemma}[Uniqueness of approximations]\label{lem:uniq-approx}
  Assume $\app(d,h)$ and $\app(d',h')$. Then
  $h(x)=h'(x)$ for all  $x\in d \cap d'$. In particular,
  \[
  h\restr(\{x\}\cup R^{-1}\{x\})=h'\restr(\{x\}\cup R^{-1}\{x\})
  \]
  for  all $x\in d \cap d'$. 
\end{lemma}

Once this is all translated to an element of \formula,
we'll be able to write the base case of the recursive definition of
$\forceisa$ using Equations~(\ref{eq:forcing-F-equ})
and~(\ref{eq:forcing-F-bel}).

@@

\section{The fundamental theorems}
These are stated as two lemmas: \emph{Truth} and
\emph{Definability}. The second one ``simply'' says that the forcing
relation is definable. The previous recursive construction amounts to
that. The Truth Lemma states that the forcing relation indeed relates
satisfaction in $M[G]$ to that in $M$. 
\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
  language of set 
  theory with all free variables displayed. Then the
  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
  \begin{enumerate}
  \item     $M[G]\models\phi(\val(\tau_1,G),\dots,\val(\tau_n,G))$ for
    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
  \item $M\models\forceisa(\phi(\tau_1,\dots,\tau_n),\PP,\leq,\1,p)$.
  \end{enumerate}
\end{lemma}
A bit more formally, one should write the parameters to the left of
$\models$:
\begin{gather*}
  \forall G : p\in G \limp M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
  \models\phi(x_1,\dots,x_n) \\
  \text{if and only if}\\
  M,[\tau_1,\dots,\tau_n,\PP,\leq,\1,p] 
  \models\forceisa(\phi(x_1,\dots,x_n),x_{n+1},x_{n+2},x_{n+3},x_{n+4})
\end{gather*}
The proof of the Truth Lemma is by recursion in elements of
$\formula$. The following auxiliary results (adapted from
\cite[IV.2.43]{kunen2011set}) are also proved by recursion in
$\formula$.
\begin{lemma}\label{lem:strengthen} 
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
  $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
  implies $M\models\forceisa(\phi,\PP,\leq,\1,p_1)$.
\end{lemma}
\begin{lemma}\label{lem:density}
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ if and only if
    $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
    dense below $p$.''
\end{lemma}
\begin{lemma}\label{lem:forcing-negation}
  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
  $M\models\forceisa(\phi,\dots)$ if and only if 
  $M\models\neg\exists
  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
\end{lemma}

\subsection{Truth for atomic formulas}
We start by proving  Lemma~\ref{lem:strengthen}.% and~\ref{lem:density}.
\begin{proof}[Proof of  Lemma~\ref{lem:strengthen} (atomic formulas)]
  Let $M$, $\lb\PP,\leq,\1\rb$, $p\in\PP$ as in the hypothesis of
  Lemma~\ref{lem:truth-lemma}, let $p_1\leq p$ and let $\phi(x_1,x_2)$
  be atomic. We 
  have to show that for every    $\tau_1,\tau_2\in M$, if
  \begin{equation} \label{eq:M-phi-atom}
    M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
    \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
  \end{equation}
  then the same holds for the environment
  $[\tau_1,\tau_2,\PP,\leq,\1,p_1]$. For this, we will deconstruct the
  definition of $\forceisa$ in several steps to pin down where the
  hypothesis of $p_1\leq p$ might be used.

  So, let $\phi$ be $x_1\in x_2$ and assume (\ref{eq:M-phi-atom}). By
  our definition of $\forceisa$  for atomic formulas, we have
  $F(\bx)=1$, where
  $\bx=\lb\equ,\tau_1,\tau_2,\PP,\leq,\1,p\rb$. Expanding the
  definition of $F$, we obtain that
  \[
  (\bx\notin A \y 1=0) \o \bigl(\bx\in A \y \exists d,h\, (\app(d,h)\y
  \bx\in d \y h(\bx)=1)\bigr)
  \]
  holds in $M$. 
  \begin{framed}
    By using appropriate absoluteness results we might translate this
    statement written as an element of $\formula$ to a sentence
    written in Isabelle/ZF first order logic, making theorems to
    simplify the statement available, as in the arguments
    that follow.
  \end{framed}
  \noindent Simplifying, we obtain 
  \begin{equation}\label{eq:1}
    \exists d,h\, (\app(d,h)\y   \bx\in d \y h(\bx)=1).
  \end{equation}
  Take such $d,h\in M$; these will be added to the environment. We
  will define $d',h'$ that satisfy 
  $\app(d',h')\y   \by\in d' \y h(\by)=1$ for
  $\by=\lb\equ,\tau_1,\tau_2,\PP,\leq,\1,p_1\rb$, thereby proving
  $F(\by)=1$ and hence the thesis.

  By  definition of $\app$ and the fact that $\bx\in d$, we have
  \[
  1= h(\bx) = G(\bx,h\restr(R^{-1}\{\bx\})).
  \]
  Now the  definition of $G$ for $\bx=\lb\equ,\dots\rb$, yields
  \[
  M,[d,h,\tau_1,\tau_2,\PP,\leq,\1,p]\models \forall
  s\in\dom(x_1)\cup\dom(x_2)\;\forall q\leq x_6 : \Phi
  \]
  with $\Phi\defi
  f(\lb\bel,s,x_1,x_{3},x_{4},x_{5},x_{6}\rb) =1  \lsii
  f(\lb\bel,s,x_2,x_{3},x_{4},x_{5},x_{6}\rb) =1$ and 
  $f\defi h\restr(R^{-1}\{\bx\})$.
  By absoluteness we have
  \[
  \forall
  \sigma\in\dom(\tau_1)\cup\dom(\tau_2)\;\forall q'\leq p :
  M,[\sigma,q',d,h,\tau_1,\tau_2,\PP,\leq,\1,p]\models  \Phi.
  \]
  Since $q'\leq p_1$ implies $q'\leq p$, we have
  \[
  \forall
  \sigma\in\dom(\tau_1)\cup\dom(\tau_2)\;\forall q'\leq p_1 :
  M,[\sigma,q',d,h,\tau_1,\tau_2,\PP,\leq,\1,p_1]\models  \Phi.
  \]
  Using absoluteness again we have
  \[
  M,[d,h,\tau_1,\tau_2,\PP,\leq,\1,p_1]\models \forall
  s\in\dom(x_1)\cup\dom(x_2)\;\forall q\leq p_1 : \Phi.
  \]
  Hence we obtain $G(\by,h\restr(R^{-1}\{\bx\}))=1$.

  Now define   $d'\defi d\cup \{\by\}$  and $h'\defi
  h\cup\{\lb\by,1\rb\}$. Since
  $R^{-1}\{\bx\} =R^{-1}\{\by\}$ and $\by\in A$, we obtain
  $\app(d',h')$. This, together with the fact that $h(\by)=1$, shows
  $F(\by)=1$ and hence (\ref{eq:M-phi-atom}).

  The proof for $\phi= x_1\in x_2$ should be similar.
\end{proof}
Lemma~\ref{lem:density} also follows the same pattern, there is a bit
more of mathematics to it. The same goes for
Lemma~\ref{lem:forcing-negation}.
